Title here

### 2020-08-06 paper/paper.Rmd
move changes made by Nicolas to Rmd

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their intial paper on projection pursuit in 1974 that "the  technique  use  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more general to help practitioner to unveil the black-box in randomised iteartive (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

In an optimization problem the goal is to find the best solution within the space of all feasible solutions which typically is represented by a set of constraints. The problem consists on optimizing an objective function $f: S \rightarrow \Re$ with $S \in \Re^n$ in a reduced spaced given by the problem constraints to either minimize or maximize a function.... Gradient based optimization has been typically used to solve such problems. However, there are many situations where derivatives of an objective function are unavailable or unreliable and therefore traditional methods based on derivatives are not the best option to solve an optimization problem.

Derivative free methods provides another option to optimise the objective function without evaluating any gradient or Hessian information and a particular class of methods: direct search, has gained its popularity through its conceptual simplicity.  However, the whole searching process in the algorithm remains a black-box. Plots are usually used to evaluate and compare the performance of different algorithms but it can easily become tedious because the code will have to be modified significantly when comparing different parameters in the algorithms. For example, a categorical variable with 5 levels can be easily mapped onto colour while mapping another categorical variable with 30 levels will not make the plot informative. Thus the plot needs to be re-designed to better suits the characteristics of the parameter (whether the parameter is a scalar or a matrix? whether the parameter is quantitative or categorical? If categorical, how many levels does the parameter have?). This motivates the  design of a visual diagnostic framework for optimisation algorithms based on the idea of a *global object*.

The remainder of the paper is organised as follows. Section \ref{DFO} gives a general overview of optimisation methods, specifically derivative free optimisation. 
Section \ref{tour} introduces the optimisation problem in a specific problem: projection pursuit guided tour and existing derivative free algorithms for solving it. 
Section \ref{vis-diag} and \ref{vis-diag-plots} together present the new construction of a visual diagnostic framework that unveil the black-box of optimisation. Section \ref{vis-diag} focuses on motivational ideas and data structure and Section \ref{vis-diag-plots} provides definitions of different diagnostic plots using grammar of graphic and examples. 
Finally, \ref{implementation} mentions the R package: ferrn that implements the visual diagnostics described above.

@@

Given an objective function $f$, one way of optimising it is to equate its gradient to zero. In modern optimisation problems, gradient information can be hard to evaluate or sometimes even impossible and Derivative-Free Optimisation (DFO) methods can be useful to approach these problems. One common class of methods in DFO is *Direct-search methods*. Coined by @hooke1961direct, direct search methods don't require any gradient or Hessian information and has gained its popularity through its simplicity in use and reliability in solving complicated practical problems. Depends on whether a random sample is used in the search, this class of methods can be further classified as *stochastic* or *deterministic*.  The stochastic version of the direct-search methods will be the main optimisation procedure analysed in this paper.

@@

[Are we using  projection pursuit/guided tour to better understand the convergence of optimization algorithms visually in combination with the algorithms discussed below? Or we are focusing on the optimisation problem only within the project pursuit context? Some of the problems listed below are also applicable to optimization problem in general too. ppp]

Below listed several issues in projection pursuit optimisation. Some are general optimisation problems, while others are more specific for PP optimisation.

- *Finding global maximum*: Although finding local maximum is relatively easy with developed algorithms, it is generally hard to guarantee global maximum in a problem where the objective function is complex or the number of decision variables is large. Also, there are discussions on how to avoid getting trapped in a local optimal in the literature.

@@

- *computation speed*: The optimisation procedure needs to be fast to compute since tours produces real-time animation of the projected data.

- *consistency result in stochastic optimisation*: In stochastic algorithm, researchers usually set a seed to ensure the algorithm producing the same result for every run. This practice supports reproducibility, while less efforts has been made to guarantee different seeds will provide the same result.

- *high-dimensional decision variable*: In projection pursuit, the decision variable includes all the entries in the projection matrix, which is high-dimensional. Researcher would be better off if they can understand the relative position of different projection matrix in the high-dimensional space.

- *role of interpolation in PP optimisation*: An optimisation procedure usually involves iteratively finding projection bases that maximises the index function, while tour requires geodesic interpolation between these bases to produce a continuous view for the users. It would be interesting to see if the interpolated bases could, in reverse, help the optimisation reach faster convergence.

@@

Coined by @friedman1974projection, projection pursuit detects interesting structures (i.e. clustering, outliers and skewness) in multivariate data via projecting onto lower dimensions. Let $\mathbf{X}_{n \times p}$ be the data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto an index value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and is used to measure the "interestingness" of a projection. A number of index functions have been proposed in the literature to detect different data structures, including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

In their initial paper, @friedman1974projection noted that "..., the technique use for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." Hence, effective optimisation algorithms are necessary for projection pursuit to find the bases that give interesting projections. While we leave the formal construction of the optimisation problem and existing algorithms in section \ref{tour-optim}, we outlines the general idea here. Given a random starting (current) basis, projection pursuit repetitively searches for candidate bases nearby until it finds one with higher index value than the current basis. In the second round, that basis becomes the current basis and the repetitive sampling  continues. The process ends until no better basis can be found or one of the termination criteria is reached.

@@

@posse1995projection proposed a random search algorithm that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$, where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{rand}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. The optimiser keeps sampling bases near the current basis until it finds one with higher index value than the current basis and then outputs it for guided tour to construct the interpolation path. A new round of search continues to find a better basis after the interpolation finishes. The halfing parameter $c$ with default value of 30 is designed to adjust the searching neighbourhood $\alpha$. When the search needs to sample more than $c$ number of basis to find an outputted basis, the neighbourhood parameter $\alpha$ will be reduced by half in the next iteration. If the optimiser can't find a better basis within the maximum number of tries $l_{\max}$, the algorithm stops. The algorithm is summarised in Algorithm \ref{random-search} for one iteration.

@@

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is sampled by $\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis is outputted. While if it is smaller, the candidate has another chance of being accepted with probability $P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}$ where $I_{\cdot}$ denotes the index value of a given basis. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial temperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Visualisation has been widely used in exploratory data analysis. Presenting information in a graphical format often allows people to see information they would otherwise not seen. This motivates us **to develop a visual framework to diagnose optimisation algorithms in projection pursuit with a aim to understand and compare different existing algorithms.** This study is meaningful because **as more complex index functions are proposed based on statistical theory, only computationally effective optimisation algorithms allow us to find the projection basis that presents interesting structure in the data**.

The idea of generalised framework for diagnostic plots is inspired by the concept of grammar of graphic [@wickham2010layered], which powers the primary graphical system in R, ggplot2 [@ggplot2]. In grammar of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by "stacked layers". Using this design, ggplot doesn't have to develop a gazillion of functions that each produces a different type of plot from a different data structure. Instead, it aesthetically maps variables (and its statistical transformation) in a dataset to different geometric objects (points, lines, box-and-whisker etc) and builds the plot through overlaying different layers.

@@

We simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution with zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distributions with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} outlines the distribution where each variable is simulated from and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to have variance of one before running the projection pursuit.

@@

We form our first dataset using variable `x1`, `x2`, `x8`, `x9` and `x10` and run the guided tour with optimiser `search_better`. When the optimisation ends, the global object will be stored and printed (it can be turned off by supplying argument `print = FALSE`). Additional messages during the optimisation can be displayed by argument `verbose = TRUE`. Below shows the first ten rows of the global object. Notice that the tibble object allows the list-column `basis` to be printed out nicely with the dimension of the projection basis readily available.

@@

A primary interest of diagnosing an optimisation algorithm is to study how it finds its optimal progressively. We could plot the index value across its natural ordering, however, different iterations may have different number of points and, towards the end of the search there could easily be hundreds of bases being tested before the target basis is found. In the plot, points from those iterations towards the end will occupy the vast majority of the plot space. This motivates to use summarisation.  Rather than knowing the index value of *every* basis, we are more interested to have a general summary of all the index value in that iteration and more importantly, the basis with the largest index value (since it prescribes the next geodesic interpolation and future searches).

Boxplot is a suitable candidate that provides five points summary of the data, however it has one drawback: it doesn't report the number of point in each box. We may risk losing information on how many points it takes to find the target basis by displaying the boxplot alone for all `tries`. Thus, the number of point in each iteration is displayed at the bottom of each box and we provide options to switch iteration with small number of points to a point geometry, which is achieved via an `cutoff` argument. A line geometry is also added to link the points with the largest index value in each iteration. This helps to visualise the improvement made in each iteration. Using the concept of *gramma of graphics* [@wickham2010layered], the plot for exploring index value can be defined in three layers as:

@@

We choose variable `x1`, `x2`, `x3`, `x8`, `x9` and `x10` to perform a 2D projection with tour. Parameter `search_f = tour::search_better` and `max.tries = 500` is used. The index value of the searching points are shown in Figure \ref{points}. Label at the bottom indicates the number of observations in each iteration and facilitates the choice of `cutoff` argument (by default `cutoff = 15`). We learn that the `search_better` quickly finds better projection basis with higher index value at first and the takes longer to find a better one later.

```{r points-tries,fig.asp=0.5, fig.cap="\\label{points}A comparison of plotting the same search points with different plot designs. The left plot doesn't efficiently use the plot space to convey information from the plot while the right plot provides good summarisation of data and number of points in each tries."}

@@

We use the same dataset as the toy example above to explore the search function `search_better` and we want to learn how the index value changes on the interpolation path for `holes` index. From the left panel of Figure \ref{interruption}, we observe that when interpolating from the current basis to the target basis, the index value may not be monotone: we could reach a basis with higher index value than the target basis on the interpolation path. In this sense, we would be better off using the basis with the highest index value on the interpolation path as the current basis for the next iteration (rather than using the target basis). Hence, an interruption is constructed to accept the interpolating bases only up to the one with the largest index value. After implementing this interruption, the search finds higher final index value with fewer steps as shown in the right panel of Figure \ref{interruption}.

```{r interruption,fig.asp=0.5, fig.cap = "\\label{interruption}Trace plots of the interpolated basis with and without the interruption. The interruption stops the interpolation when the index value starts to decrease at id = 60. The implementation of the interuption finds an ending basis with higher index value using fewer steps. "}
p1_anno <- interrupt_no %>% filter(info == "interpolation") %>% mutate(id = row_number()) %>% filter(id %in% c(44, 60, 62)) %>% 
  mutate(anno = c("current basis", "interpolated basis", "target basis"))

@@

1) the chosen basis can't be too close to the current basis 
2) the percentage improvement of the index value can't be too small 
3) the searching neighbourhood can't be too small
4) the number of iteration can't exceed the `max.tries`

@@

```{r polish, fig.asp=0.5,fig.cap="\\label{trace-compare}Breakdown of index value when using different max.tries in search better in conjunction with search polish. Both attain the same final index value after the polishing while using a max.tries 25 is not sufficient to find the ture maximum."}

@@

*Example: understand search_geodesic* [feel like this example is merely explaining search geodesic algorithm, so maybe introduce the animated plot here? xxx] `search_geodesic` is a two-stage ascending algorithm with four different stages in the search and a PCA plot useful to understand how the algorithm progresses and the relative position of each basis in the PCA projected 2D space. Starting from the start basis, a directional search is conducted in a narrow neighbourhood on five random directions. The best one is picked and a line search is then run on the geodesic direction to find the target basis. The starting and target bases are then interpolated. In the next iteration, the target basis becomes the current basis and then procedures continues.

@@

*Example: initial value for polishing alpha* `search_polish` is a brutal-force algorithm that evaluate 1000 points in the neighbourhood at each loop. Setting an appropriate initial value for polish_alpha would avoid wasting search on large vector space that are not likely to produce higher index value. The default initial value for polishing step is 0.5 and we are interested in whether this is an appropriate initial value to use after `search_geodesic`. The problem is a 1D projection of the small dataset using `search_geodesic` and followed by `search_polish`. The top-left panel of Figure \ref{polish-alpha} displays all the projection bases on the first two principal components, coloured by the `polish_alpha`. We can observe that rather than concentrating on the ending basis from `search_geodesic` as what polishing step is designed, `search_polish` searches a much larger vector space, which is unnecessary. Thus a customised smaller initial value for `polish_alpha` would be ideal. One way to do this is to initialised `polish_alpha` as the projection distance between the last two target bases. The top-right panel of Figure \ref{polish-alpha} shows a more desirable concentrated searching space near the ending basis. Both specifications of initial value allow the searches to reach the same ending index values.

@@

Now we enlarge the dataset to include two informative variables: `x2` and `x3` and remain 1D projection. In this case, two local maximum appear when projection matrix being $[0, 1, 0, 0, 0, 0]$ and $[0, 0, 1 ,0, 0, 0]$.

Using different seeds in `search_better` allows us to find both local maximum as in Figure \ref{1d-2var-different-seeds}. Comparing the maximum of both, we can see that the global maximum happens when `x2` is found. It is natural to ask then if there is an algorithm that can find the global maximum without trying on different seeds? `search_better_random` manages to do it via a metropolis-hasting random search as shown in Figure \ref{1d-2var-better-random}, although at a higher cost of number of points to evaluate.

@@

We can also plot the searching points of all the three algorithms in the searching space and explore their relative position against each other using principal components. As shown in Figure \ref{1d-2var-explore-proj-pca}, the bases from better1 and better2 only search a proportion of the searching space while better_random produces a more exhaustive search. The large overlapping of better1 and better_random is explained by the fact that both algorithms finds x2 in the end.

>>>>

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

In an optimization problem the goal is to find the best solution within the space of all feasible solutions which typically is represented by a set of constraints. The problem consists in optimizing an objective function $f: S \rightarrow \Re$ with $S \in \Re^n$ in a reduced spaced given by the problem constraints to either minimize or maximize a function.... Gradient based optimization has been typically used to solve such problems. However, there are many situations where derivatives of an objective function are unavailable or unreliable and therefore traditional methods based on derivatives are not the best option to solve an optimization problem.

Derivative free methods provide another option to optimise the objective function without evaluating any gradient or Hessian information and a particular class of methods, direct search, has gained popularity through its conceptual simplicity.  However, the whole searching process in the algorithm remains a black-box. Plots are usually used to evaluate and compare the performance of different algorithms but it can easily become tedious because the code will have to be modified significantly when comparing different parameters in the algorithms. For example, a categorical variable with 5 levels can be easily mapped onto colour while mapping another categorical variable with 30 levels will not make the plot informative. Thus the plot needs to be re-designed to better suit the characteristics of the parameter (whether the parameter is a scalar or a matrix? whether the parameter is quantitative or categorical? If categorical, how many levels does the parameter have?). This motivates the  design of a visual diagnostic framework for optimisation algorithms based on the idea of a *global object*.

The remainder of the paper is organised as follows. Section \ref{DFO} gives a general overview of optimisation methods, specifically derivative free optimisation. 
Section \ref{tour} introduces the optimisation problem in a specific problem: projection pursuit guided tour and existing derivative free algorithms for solving it. 
Section \ref{vis-diag} and \ref{vis-diag-plots} together present the new construction of a visual diagnostic framework that unveil the black-box of optimisation. Section \ref{vis-diag} focuses on motivational ideas and data structure and Section \ref{vis-diag-plots} provides definitions of different diagnostic plots using grammar of graphic and examples. 
Finally, Section \ref{implementation} mentions the R package: ferrn that implements the visual diagnostics described above.

@@

Given an objective function $f$, one way of optimising it is to equate its gradient to zero. In modern optimisation problems, gradient information can be hard to evaluate or sometimes even impossible and Derivative-Free Optimisation (DFO) methods can be useful to approach these problems. One common class of methods in DFO is *Direct-search methods*. Coined by @hooke1961direct, direct search methods don't require any gradient or Hessian information and has gained its popularity through its simplicity in use and reliability in solving complicated practical problems. Depending on whether a random sample is used in the search, this class of methods can be further classified as *stochastic* or *deterministic*.  The stochastic version of the direct-search methods will be the main optimisation procedure analysed in this paper.

@@

[Are we using  projection pursuit/guided tour to better understand the convergence of optimization algorithms visually in combination with the algorithms discussed below? Or we are focusing on the optimisation problem only within the projection pursuit context? Some of the problems listed below are also applicable to optimization problem in general too. ppp]

Below are listed several issues in projection pursuit optimisation. Some are general optimisation problems, while others are more specific too PP optimisation.

- *Finding global maximum*: Although finding local maximum is relatively easy with developed algorithms, it is generally hard to guarantee global maximum in a problem where the objective function is complex or the number of decision variables is large. Also, there are discussions on how to avoid getting trapped in a local optimum in the literature.

@@

- *computation speed*: The optimisation procedure needs to be fast to compute since the tour produces real-time animation of the projected data.

- *consistency result in stochastic optimisation*: In stochastic algorithm, researchers usually set a seed to ensure the algorithm produces the same result for every run. This practice supports reproducibility, while less efforts has been made to guarantee different seeds will provide the same result.

- *high-dimensional decision variable*: In projection pursuit, the decision variable includes all the entries in the projection matrix, which is high-dimensional. Researchers would be better off if they could understand the relative position of different projection matrix in the high-dimensional space.

- *role of interpolation in PP optimisation*: An optimisation procedure usually involves iteratively finding projection bases that maximise the index function, while tour requires geodesic interpolation between these bases to produce a continuous view for the users. It would be interesting to see if the interpolated bases could, in reverse, help the optimisation reach faster convergence.

@@

Coined by @friedman1974projection, projection pursuit detects interesting structures (i.e. clustering, outliers and skewness) in multivariate data via projecting onto lower dimensions. Let $\mathbf{X}_{n \times p}$ be the data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ to be an index function that maps the projection basis $\mathbf{A}$ onto an index value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and is used to measure the "interestingness" of a projection. A number of index functions have been proposed in the literature to detect different data structures, including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

In their initial paper, @friedman1974projection noted that "..., the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." Hence, effective optimisation algorithms are necessary for projection pursuit to find the bases that give interesting projections. While we leave the formal construction of the optimisation problem and existing algorithms to section \ref{tour-optim}, we outline the general idea here. Given a random starting (current) basis, projection pursuit repeatedly searches for candidate bases nearby until it finds one with higher index value than the current basis. In the second round, that basis becomes the current basis and the repetitive sampling  continues. The process ends until no better basis can be found or one of the termination criteria is reached.

@@

@posse1995projection proposed a random search algorithm that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$, where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{rand}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. The optimiser keeps sampling bases near the current basis until it finds one with higher index value than the current basis and then outputs it for guided tour to construct the interpolation path. A new round of search continues to find a better basis after the interpolation finishes. The halving parameter $c$ with default value of 30 is designed to adjust the searching neighbourhood $\alpha$. When the search needs to sample more than $c$ number of basis to find an outputted basis, the neighbourhood parameter $\alpha$ will be reduced by half in the next iteration. If the optimiser can't find a better basis within the maximum number of tries $l_{\max}$, the algorithm stops. The algorithm is summarised in Algorithm \ref{random-search} for one iteration.

@@

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{\text{rand}}$, a candidate basis is sampled by $\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{\text{rand}}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis is outputted. While if it is smaller, the candidate has another chance of being accepted with probability $P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}$ where $I_{\cdot}$ denotes the index value of a given basis. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial temperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
        \If{$P > U$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Visualisation has been widely used in exploratory data analysis. Presenting information in a graphical format often allows people to see information they would otherwise not see. This motivates us **to develop a visual framework to diagnose optimisation algorithms in projection pursuit with the aim to understand and compare different existing algorithms.** This study is meaningful because **as more complex index functions are proposed based on statistical theory, only computationally effective optimisation algorithms allow us to find the projection basis that presents interesting structure in the data**.

The idea of generalised framework for diagnostic plots is inspired by the concept of grammar of graphic [@wickham2010layered], which powers the primary graphical system in R, ggplot2 [@ggplot2]. In grammar of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by "stacked layers". Using this design, ggplot does not have to develop a gazillion of functions that each produces a different type of plot from a different data structure. Instead, it aesthetically maps variables (and its statistical transformation) in a dataset to different geometric objects (points, lines, box-and-whisker etc) and builds the plot through overlaying different layers.

@@

We simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution with zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distributions with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} outlines the distribution where each variable is simulated from and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to have unit variance before running the projection pursuit.

@@

We form our first dataset using variables `x1`, `x2`, `x8`, `x9` and `x10` and run the guided tour with optimiser `search_better`. When the optimisation ends, the global object will be stored and printed (it can be turned off by supplying argument `print = FALSE`). Additional messages during the optimisation can be displayed by argument `verbose = TRUE`. Below shows the first ten rows of the global object. Notice that the tibble object allows the list-column `basis` to be printed out nicely with the dimension of the projection basis readily available.

@@

A primary interest of diagnosing an optimisation algorithm is to study how it finds its optimum progressively. We could plot the index value across its natural ordering, however, different iterations may have different number of points and, towards the end of the search there could easily be hundreds of bases being tested before the target basis is found. In the plot, points from those iterations towards the end will occupy the vast majority of the plot space. This motivates to use summarisation.  Rather than knowing the index value of *every* basis, we are more interested to have a general summary of all the index value in that iteration and more importantly, the basis with the largest index value (since it prescribes the next geodesic interpolation and future searches).

Boxplot is a suitable candidate that provides five points summary of the data, however it has one drawback: it does not report the number of points in each box. We may risk losing information on how many points it takes to find the target basis by displaying the boxplot alone for all `tries`. Thus, the number of point in each iteration is displayed at the bottom of each box and we provide options to switch iteration with small number of points to a point geometry, which is achieved via a `cutoff` argument. A line geometry is also added to link the points with the largest index value in each iteration. This helps to visualise the improvement made in each iteration. Using the concept of *gramma of graphics* [@wickham2010layered], the plot for exploring index value can be defined in three layers as:

@@

We choose variables `x1`, `x2`, `x3`, `x8`, `x9` and `x10` to perform a 2D projection with tour. Parameter `search_f = tour::search_better` and `max.tries = 500` is used. The index value of the searching points are shown in Figure \ref{points}. The label at the bottom indicates the number of observations in each iteration and facilitates the choice of `cutoff` argument (by default `cutoff = 15`). We learn that the `search_better` quickly finds better projection basis with higher index value at first and then takes longer to find a better one later.

```{r points-tries,fig.asp=0.5, fig.cap="\\label{points}A comparison of plotting the same search points with different plot designs. The left plot does not use the plot space efficiently to convey information from the plot while the right plot provides good summarisation of data and number of points in each tries."}

@@

We use the same dataset as the toy example above to explore the search function `search_better` and we want to learn how the index value changes on the interpolation path for the `holes` index. From the left panel of Figure \ref{interruption}, we observe that when interpolating from the current basis to the target basis, the index value may not be monotone: we could reach a basis with a higher index value than the target basis on the interpolation path. In this sense, we would be better off using the basis with the highest index value on the interpolation path as the current basis for the next iteration (rather than using the target basis). Hence, an interruption is constructed to accept the interpolating bases only up to the one with the largest index value. After implementing this interruption, the search finds higher final index value with fewer steps as shown in the right panel of Figure \ref{interruption}.

```{r interruption,fig.asp=0.5, fig.cap = "\\label{interruption}Trace plots of the interpolated basis with and without interruption. The interruption stops the interpolation when the index value starts to decrease at id = 60. The implementation of the interuption finds an ending basis with higher index value using fewer steps. "}
p1_anno <- interrupt_no %>% filter(info == "interpolation") %>% mutate(id = row_number()) %>% filter(id %in% c(44, 60, 62)) %>% 
  mutate(anno = c("current basis", "interpolated basis", "target basis"))

@@

1) the chosen basis can't be too close to the current basis 
2) the percentage improvement of the index value can't be too small 
3) the searching neighbourhood can't be too small
4) the number of iteration can't exceed `max.tries`

@@

```{r polish, fig.asp=0.5,fig.cap="\\label{trace-compare}Breakdown of index value when using different max.tries in search better in conjunction with search polish. Both attain the same final index value after the polishing while using a `max.tries` of 25 is not sufficient to find the true maximum."}

@@

*Example: understand search_geodesic* [feel like this example is merely explaining search geodesic algorithm, so maybe introduce the animated plot here? xxx] `search_geodesic` is a two-stage ascending algorithm with four different stages in the search and a PCA plot useful to understand how the algorithm progresses and the relative position of each basis in the PCA projected 2D space. Starting from the start basis, a directional search is conducted in a narrow neighbourhood on five random directions. The best one is picked and a line search is then run on the geodesic direction to find the target basis. The starting and target bases are then interpolated. In the next iteration, the target basis becomes the current basis and then the procedures continues.

@@

*Example: initial value for polishing alpha* `search_polish` is a brute-force algorithm that evaluate 1000 points in the neighbourhood at each loop. Setting an appropriate initial value for polish_alpha would avoid wasting search on large vector space that are not likely to produce higher index value. The default initial value for polishing step is 0.5 and we are interested in whether this is an appropriate initial value to use after `search_geodesic`. The problem is a 1D projection of the small dataset using `search_geodesic` and followed by `search_polish`. The top-left panel of Figure \ref{polish-alpha} displays all the projection bases on the first two principal components, coloured by the `polish_alpha`. We can observe that rather than concentrating on the ending basis from `search_geodesic` as what polishing step is designed, `search_polish` searches a much larger vector space, which is unnecessary. Thus a customised smaller initial value for `polish_alpha` would be ideal. One way to do this is to initialised `polish_alpha` as the projection distance between the last two target bases. The top-right panel of Figure \ref{polish-alpha} shows a more desirable concentrated searching space near the ending basis. Both specifications of initial value allow the searches to reach the same ending index values.

@@

Now we enlarge the dataset to include two informative variables: `x2` and `x3` and remain 1D projection. In this case, two local maxima appear with projection matrix being $[0, 1, 0, 0, 0, 0]$ and $[0, 0, 1 ,0, 0, 0]$.

Using different seeds in `search_better` allows us to find both local maxima d as in Figure \ref{1d-2var-different-seeds}. Comparing the maximum of both, we can see that the global maximum happens when `x2` is found. It is natural to ask then if there is an algorithm that can find the global maximum without trying on different seeds? `search_better_random` manages to do it via a Metropolis-hasting random search as shown in Figure \ref{1d-2var-better-random}, although at a higher cost of number of points to evaluate.

@@

We can also plot the searching points of all three algorithms in the searching space and explore their relative position against each other using principal components. As shown in Figure \ref{1d-2var-explore-proj-pca}, the bases from better1 and better2 only search a proportion of the searching space while better_random produces a more exhaustive search. The large overlapping of better1 and better_random is explained by the fact that both algorithms find x2 in the end.
